---
title: "Figures Included in the Botanical Homesis Testing Manuscript"
author: "Ying Jin"
date: "`r Sys.Date()`"
output: 
  html_document:
    number_section: true
    toc: true
    toc_depth: 3
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, dev = "pdf", fig.path = "ReportFigs/")

set.seed(333)

library(tidyverse)
theme_set(theme_minimal())
library(here)
library(RColorBrewer)
library(scales)
library(SRMERS)
# Pick a Brewer palette
my_cols <- c(brewer.pal(8, "Set2")[c(2,6,3,8)], "#D3D3D3")
# Apply globally
options(
  ggplot2.discrete.colour = my_cols,
  ggplot2.discrete.fill   = my_cols
)

source("../R/SharpScatter.R") # plot function

# chemical information
sample_info <- read_tsv(here("Data/botanical_sample_info.tsv")) 

# fix duplicate sample name
dup_chems <- sample_info %>% group_by(sample_name) %>% summarise(n= n()) %>% 
  filter(n>1)
sample_info <- sample_info %>% 
  mutate(sample_name = ifelse(sample_name%in% dup_chems$sample_name, 
                              paste0(sample_name, " (lot ", sample_lot, ")"),
                              sample_name))

# raw dose-response data
ds_curves <- read_tsv("../Data/botanical_norm_resp.tsv")
ds_curves <- ds_curves %>%
  rename(readout=protocol_readout) %>%
  left_join(sample_info, by = "sample_id") %>%
  mutate(concs=log(concs))
```


```{r load_output}
# load p values from 21 iterations
output_names <- list.files("../Data/VoteOutput")
shapes <- c("increase", "decrease", "convex", "concave")

## containers
final_pvals <- list()
final_shape <- list()

for(k in seq_along(output_names)){
  load(here(paste0("Data/VoteOutput/", output_names[k])))
  # median p-values
  df_medp <- output_chem_k %>% group_by(Readout) %>% summarise_at(shapes, median)
  df_medp <- df_medp %>% mutate(chemical = substr(output_names[k], 1, 15))
  final_pvals[[k]] <- df_medp
  
  # shape
  df_shape <- df_medp %>%
    mutate_at(shapes, function(x){x<0.05})
  df_shape$shape <- apply(df_shape[, shapes], 1, 
                          function(x){paste(shapes[x], collapse = "-")})
  final_shape[[k]] <- df_shape
  
}
```


```{r pvalues_clean}
# table of p values
df_pvals <- bind_rows(final_pvals) %>%
  rename(sample_id = chemical) %>% 
  left_join(sample_info, by = "sample_id") 

# write.csv(df_pvals, file = here("Data/allPvals.csv"), row.names = F)

# remove channel 1 and 2
df_pvals <- df_pvals %>%  
  filter(!str_ends(Readout, "ch1") & !str_ends(Readout, "ch2"))
# length(unique(df_pvals$Readout))
# length(unique(df_pvals$sample_id))
```

```{r shape_clean}
df_shape <-bind_rows(final_shape)
# shape category
df_shape <- df_shape %>% mutate(
  shape2 = fct_collapse(shape,
                        "Hormesis" = c("decrease-concave", "increase-convex"), 
                        "Monotone" = c("decrease", "increase"), 
                        "U shape" = c("convex", "concave"),
                        "No shape" = c(""),
                        other_level = "Uncertain")) %>% 
  mutate(shape2 = fct_relevel(shape2, "Hormesis", after = 0)) %>% 
  mutate(shape2 = fct_relevel(shape2, "No shape", after = Inf))

# remove channel 1 and 2
df_shape <- df_shape %>%
  filter(!str_ends(Readout, "ch1") & !str_ends(Readout, "ch2"))
# length(unique(df_shape$chemical)) # 90 chemicals
# length(unique(df_shape$Readout)) # 54 readouts

# add sample info
df_shape <- df_shape %>% 
  rename(sample_id = chemical) %>%
  left_join(sample_info)
```

# Overall summarization

```{r overall_sum, fig.height=20, fig.width=15}
# counts
table(df_shape$shape2)
```



```{r overall_sum_label, fig.height=20, fig.width=15}
# plot by botanical group
df_shape %>%
  # group assays by channel 
  mutate(out_ch = sub(".*_", "", Readout), .after = "Readout") %>%
  mutate(out_ch = ifelse(str_ends(out_ch, "h"), "time", out_ch)) %>%
  ggplot(aes(x=Readout, y = sample_id, fill=shape2))+
  geom_tile()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 10),
        axis.text = element_text(size = 10),
        # strip.text.y = element_text(angle = 0, size = 20), 
        strip.text.y = element_text(angle = 0, size = 10),
        strip.text.x = element_text(size = 10),
        plot.margin = margin(10, 20, 10, 20),
        legend.title = element_blank(),
        legend.text = element_text(size = 10),
        legend.position = "top")+
  scale_y_discrete(breaks = df_shape$sample_id, labels = df_shape$sample_name)+
  scale_fill_manual(values = my_cols)+
  facet_grid(rows = vars(str_wrap(botanical_group)), cols = vars(out_ch), 
             scales = "free", space = "free")+
  labs(x=" ", y= " ")
```


## Scatterplot example of one assay

```{r one_assay_scatter_exp, fig.height=8, fig.width=10}
df_plot <- df_pvals %>% filter(Readout == "mixture-are-bla-p1_ratio") %>%
  left_join(sample_info)

SharpScatter(df_plot$increase, df_plot$decrease, df_plot$concave, df_plot$convex, 
             col = df_plot$botanical_group)+
  labs(col="", title = "mixture-are-bla-p1_ratio")
```
```{r, eval = F, fig.height=8, fig.width=10}

# for supplements, scatter plots for all assays
all_assays <- unique(df_pvals$Readout)

for(r in all_assays){
  df_plot_r <- df_pvals %>% 
    filter(Readout ==r) %>%
    left_join(sample_info)
  scatter_r <- SharpScatter(df_plot_r$increase, 
                            df_plot_r$decrease, 
                            df_plot_r$concave, 
                            df_plot_r$convex, 
             col = df_plot_r$botanical_group)+
  labs(col="", title = r)
  
  filepath <- paste0("Manuscripts/ReportFigs/ScatterAssay/", r, ".jpeg")
  ggsave(here(filepath), height = 8, width = 10)
  
}
```

## Table summary by botanical group (chemical) 

```{r}
# hormesis summary table by chemical groups
# table(df_shape$shape2)
df_shape %>% select(botanical_group, sample_name, Readout, shape2) %>% 
  group_by(botanical_group) %>% 
  summarise(nhorm = sum(shape2=="Hormesis"),
            ncurve = n(),
            pct = round(100*(nhorm/ncurve), 2)) %>%
  mutate('N (pct)' = paste0(nhorm, " (", pct, "%)")) %>%
  arrange(desc(nhorm)) %>%
  select(botanical_group, 'N (pct)') %>%
  rename("Botanical Group"=botanical_group) %>%
  kableExtra::kable(align = "lc",
                    caption = "Number of dose-response curves identified as hormesis, and percentage of hormesis curves out of all dose-reponse curves in each botanical group.") %>% 
  kableExtra::kable_styling(full_width = F)
  
```

## Table summary by assays

```{r}
# ds_horm %>% distinct(.)
# correspond each assay with their readouts
channel_assay_list <- df_shape %>%
  filter(shape2=="Hormesis") %>%
  select(Readout) %>%
  distinct(.) %>%
  separate(Readout, into = c("Assay", "Channel"), sep = "_", remove = F) %>%
  group_by(Channel)
channel_assay_list <- split(channel_assay_list$Assay, f = channel_assay_list$Channel)

df_shape <- df_shape %>% 
  separate(Readout, into = c("Assay", "Channel"), sep = "_", remove = F)
```


```{r, message=FALSE}
# luc and viability 
df_shape %>% group_by(Assay, Channel) %>% 
  filter(Assay %in% channel_assay_list$luc & Assay %in% channel_assay_list$via) %>%
  group_by(Assay, Channel) %>%
  summarise(nhorm = sum(shape2=="Hormesis"),
            ncurve = n(),
            pct = round(100*(nhorm/ncurve), 2)) %>%
  mutate('N (pct)' = paste0(nhorm, " (", pct, "%)")) %>%
  select(Assay, Channel, 'N (pct)') %>%
  pivot_wider(id_cols = "Assay", values_from ='N (pct)', names_from = "Channel") %>% 
  knitr::kable(caption = "Number of dose-response curves identified as hormesis, and percentage of hormesis curves out of all dose-reponse curves of each assay in each channel.") %>%
  kableExtra::kable_styling(full_width = F)
  
```
```{r, message=FALSE}
# ratio and viability
df_shape %>% 
  filter(Assay %in% channel_assay_list$ratio & Assay %in% channel_assay_list$via) %>%
  group_by(Assay, Channel) %>%
  summarise(nhorm = sum(shape2=="Hormesis"),
            ncurve = n(),
            pct = round(100*(nhorm/ncurve), 2)) %>%
  mutate('N (pct)' = paste0(nhorm, " (", pct, "%)")) %>%
  select(Assay, Channel, 'N (pct)') %>%
  pivot_wider(id_cols = "Assay", values_from ='N (pct)', names_from = "Channel") %>% 
  knitr::kable(caption = "Number of dose-response curves identified as hormesis, and percentage of hormesis curves out of all dose-reponse curves of each assay in each channel.") %>%
  kableExtra::kable_styling(full_width = F)
```

```{r, message=FALSE}
# time
df_shape %>% group_by(Assay, Channel) %>% 
  filter(Assay %in% channel_assay_list$`0h`) %>%
  group_by(Assay, Channel) %>%
  summarise(nhorm = sum(shape2=="Hormesis"),
            ncurve = n(),
            pct = round(100*(nhorm/ncurve), 2)) %>%
  mutate('N (pct)' = paste0(nhorm, " (", pct, "%)")) %>%
  select(Assay, Channel, 'N (pct)') %>%
  pivot_wider(id_cols = "Assay", values_from ='N (pct)', names_from = "Channel") %>% 
  knitr::kable(caption = "Number of dose-response curves identified as hormesis, and percentage of hormesis curves out of all dose-reponse curves of each assay at each time.") %>%
  kableExtra::kable_styling(full_width = F)
```


# Individual dose-response curves examples

- Hormesis: NCGC0037

```{r shape_exp, fig.height=6, fig.width=7}
data.frame(
sample_id = c("NCGC00372845-01", "NCGC00372744-01", "NCGC00372687-01", "NCGC00372775-01"),
readout = c("mixture-rt-viability-hek293-glo-p1_8h", "mixture-pparg-bla-antagonist-p1_ratio",
            "mixture-rt-viability-hek293-glo-p1_8h", "mixture-rt-viability-hek293-flor-p1_40h")
) %>%
  mutate(shape = factor(c("(a) Inverted J-shape", "(b) Monotone", "(c) Inverted U-shape", "(d) No shape"))) %>%
  left_join(ds_curves, by = c("sample_id", "readout")) %>%
  ggplot(aes(x=concs, y=resps))+
  geom_smooth()+
  geom_point(size = 1)+
  facet_wrap(~shape, scales="free")+
  labs(x = "Dose", y = "Response")
  


```


# Demonstration of how to use the scatterplot

## Subspace labels

```{r}
# add shading
alpha <- 0.05

# boundaries
cuts <- c(-1, -(1 - alpha), (1 - alpha), 1)

# define intervals (3 per axis)
x_intervals <- data.frame(
  xmin = cuts[-length(cuts)],
  xmax = cuts[-1]
)

y_intervals <- data.frame(
  ymin = cuts[-length(cuts)],
  ymax = cuts[-1]
)

# now combine intervals → 3 × 3 = 9 rectangles
rects <- merge(x_intervals, y_intervals)

# optional: assign colors (9 distinct ones)
rects$fill <- c("Uncertain", "Monotone", "Hormesis", 
                    "U shape", "No shape", "U shape", 
                    "Hormesis", "Monotone", "Uncertain")
rects$fill <- factor(rects$fill, levels = c("Hormesis", "U shape", "Monotone", "Uncertain","No shape"))
```

```{r}
alpha = 0.05
# scaling functions
NewScale <- function(x){
      newx <- ifelse(abs(x)>(1-alpha), 1-((1-abs(x))/alpha)*0.5, abs(x)*0.5/(1-alpha))
      newx = sign(x)*newx
      return(newx)
    }

InverseNewScale <- function(y){
  origy <- ifelse(abs(y) > 0.5, 1 - ((1 - abs(y)) * alpha / 0.5), abs(y) * (1-alpha) / 0.5)
  origy = sign(y) * origy
  return(origy)
}
# transformation object
NewTrans <- trans_new(
  name = "NewTrans",
  transform = NewScale,
  inverse = InverseNewScale
)
```


```{r subspace}
ggplot()+
    geom_hline(yintercept = 0, linewidth = 0.5) +  # center y-axis
    geom_vline(xintercept = 0, linewidth = 0.5) +  # center x-axis
    labs(x=" ", y=" ")+
    theme(panel.grid = element_blank(), axis.text = element_blank(), axis.ticks = element_blank())+
    annotate("text", x = c(-0.9, 0.9), y = c(0,0), label = c("Convex", "Concave"), vjust = 1.1)+
    annotate("text", y = c(-1, 1), x = c(0,0), label = c("Decrease", "Increase"), hjust = 1.1)+
    theme_minimal()+
    geom_hline(yintercept = c(1-0.05, -1+0.05), linetype = "dashed", linewidth = 0.5)+
    geom_vline(xintercept = c(1-0.05, -1+0.05), linetype = "dashed", linewidth = 0.5)+
    scale_x_continuous(trans = NewTrans, limits = c(-1, 1))+
    scale_y_continuous(trans = NewTrans, limits = c(-1, 1))+
    geom_rect(
        data = rects,
        aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = fill),
        alpha = 0.8)+
  scale_fill_manual(values = my_cols)

```


## Chemical: Turmeric (lot 90H73) (NCGC00372872-01)

- ambigous: aromatase-er-aromatase-p1_luc
- no shape: elg1-luc-agonist-p1_luc
- hormesis, convex-increase: rt-viability-hek293-flor-p1_0h
- hormesis, concave-decrease: rt-viability-hepg2-glo-p1_0h
- U shape, concave: rt-viability-hek293-glo-p1_0h
- U shape, convex: 
- Monotone, increase: 
- Monoton decrease: mmp-p1_ratio

```{r scatter_demo_A, fig.height=8, fig.width=10}
load(here("Data/VoteOutput/NCGC00372872-01.RData"))
# examination
# plot_outs <- df_shape %>% filter(sample_id == "NCGC00372872-01" & shape2 == "U shape")
# plot_outs <- plot_outs$Readout
plot_outs <- c("aromatase-er-aromatase-p1_luc", "elg1-luc-agonist-p1_luc",
               "rt-viability-hek293-flor-p1_0h" ,"rt-viability-hepg2-glo-p1_0h",
               "rt-viability-hek293-glo-p1_0h", "mmp-p1_ratio")
plot_outs <- paste0("mixture-", plot_outs)
# color_label <- c("Ambiguous", "No shape", 
#                  "Hormesis (convex-increase) ", "Hormesis (concave-decrease)",
#                  "U shape (concave) & ambiguous", 
#                  "Monotone (decrease)")
# names(color_label) <- plot_outs

df_plot <- output_chem_k %>%
  filter(Readout %in% plot_outs) 

SharpScatter(df_plot$increase, df_plot$decrease,
             df_plot$concave,df_plot$convex, 
             col = df_plot$Readout, size_point = 2)+
  # scale_color_discrete("", labels = color_label)+
  theme(legend.direction = "horizontal", legend.position = "top")+
  labs(col=" ")
```

```{r scatter_demo_B, fig.height=6, fig.width=10}
ds_curves %>%
  filter(sample_id=="NCGC00372872-01" & readout %in% plot_outs) %>%
    ggplot(aes(x=concs, y=resps))+
    geom_point(aes(col = readout))+
    geom_smooth(aes(col = readout), span = 0.7)+
    theme(legend.position = "none")+
    facet_wrap(~readout, scales="free", nrow = 2)+
  labs(x="Dose", y = "Response")

```



# Examples of false negative due to insufficient dose grid

```{r}
source("../R/SHARPtest.R")

# true
x <- seq(0, 1, by = 0.01)
fx <- -(sqrt(x)+0.2)*(1-x)^2+1
df_tr <- data.frame(x, fx)
# plot(x, fx)
```



```{r}
# observation 1
xobs <- seq(0, 0.4, length.out = 15)
#y <- xobs+sin(xobs)+rnorm(10, 0, 0.1)
y <-  -(sqrt(xobs)+0.2)*(1-xobs)^2+1+rnorm(length(xobs), 0, 0.01)
df_obs <- data.frame(xobs, y)

pvals <- SHARPtest(df_obs, mixed = F, xName = "xobs", yName = "y")

# format p values
pvals <- pvals[c("increase", "decrease", "convex", "concave")]
pval_txt <- scales::pvalue(as_vector(pvals), add_p = T)
pval_txt <- paste(names(pval_txt), pval_txt, sep = ": ", collapse = "\n")

# fn_p1 <- ggplot()+
#   geom_line(data = df_tr, aes(x=x, y=fx))+
#   geom_point(data=df_obs, aes(x=xobs, y=y))+
#   annotate("text", x = 0.9, y = 0.6, label = pval_txt)+
#   labs(x="Dose", y="Response", title = "Undetected hormesis: U-shape")
```

```{r}
# observation 2
# monotone
xobs2 <- seq(0.5, 0.8, length.out = 15)
y2 <-  -(sqrt(xobs2)+0.2)*(1-xobs2)^2+1+rnorm(length(xobs2), 0, 0.01)
df_obs2 <- data.frame(xobs2, y2)
# plot(df_obs2$xobs2, df_obs2$y2)

pvals2 <- SHARPtest(df_obs2, mixed = F, xName = "xobs2", yName = "y2")

# format p values
pvals2 <- pvals2[c("increase", "decrease", "convex", "concave")]
pval_txt2 <- scales::pvalue(as_vector(pvals2), add_p = T)
pval_txt2 <- paste(names(pval_txt2), pval_txt2, sep = ": ", collapse = "\n")

# fn_p2 <- ggplot()+
#   geom_line(data = df_tr, aes(x=x, y=fx))+
#   geom_point(data=df_obs2, aes(x=xobs2, y=y2))+
#   annotate("text", x = 0.9, y = 0.6, label = pval_txt2)+
#   labs(x="Dose", y="Response", title = "Undetected hormesis: Monotone")

```

```{r false_negative, fig.height=5, fig.width=8}
# ggpubr::ggarrange(fn_p1, fn_p2, nrow = 1)

ggplot()+
  geom_line(data = df_tr, aes(x=x, y=fx))+
  geom_point(data=df_obs, aes(x=xobs, y=y, col = "U shape"))+
  geom_point(data=df_obs2, aes(x=xobs2, y=y2, col = "Monotone"))+
  annotate("text", x = 0.1, y = 0.95, label = pval_txt, col = "#FC8D62")+
  annotate("text", x = 0.9, y = 0.6, label = pval_txt2, col = "#8DA0CB")+
  labs(x="Dose", y="Response", title = "Undetected hormesis", col = " ")+
  scale_color_manual(values = c("U shape" = "#FC8D62", 
                                "Monotone" = "#8DA0CB"))
```
