---
title: "Repeated SHARP Shape Detection"
author: "Ying Jin"
date: "`r Sys.Date()`"
output: 
  html_document:
    number_section: true
    toc: true
    toc_depth: 3
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, dev = "jpeg", fig.path = "ReportFigs/")


library(tidyverse)
theme_set(theme_minimal())
library(here)
library(RColorBrewer)
# Pick a Brewer palette
my_cols <- c(brewer.pal(8, "Set2")[c(2,6,3,8)], "#D3D3D3")
# Apply globally
options(
  ggplot2.discrete.colour = my_cols,
  ggplot2.discrete.fill   = my_cols
)

source("../R/SharpScatter.R")

sample_info <- read_tsv(here("Data/botanical_sample_info.tsv"))

# fix duplicate sample name
dup_chems <- sample_info %>% group_by(sample_name) %>% summarise(n= n()) %>% 
  filter(n>1)
sample_info <- sample_info %>% 
  mutate(sample_name = ifelse(sample_name%in% dup_chems$sample_name, 
                              paste0(sample_name, " (lot ", sample_lot, ")"),
                              sample_name))

# table(df_shape$shape2)
ds_curves <- read_tsv("../Data/botanical_norm_resp.tsv")
ds_curves <- ds_curves %>%
  rename(readout=protocol_readout) %>%
  mutate(readout=gsub("mixture-","", readout)) %>%
  left_join(sample_info, by = "sample_id") %>%
  mutate(concs=log(concs))
```


# Raw curves 

```{r exp_raw_curves, fig.width=12, fig.height=5}
# rand_readout <- sample(unique(ds_curves$readout), 1)
chems <- c("NCGC00372678-01", "NCGC00372775-01", "NCGC00372575-01")
outs <- c("hse-bla-p1_via", "er-luc-bg1-4e2-agonist-p1_via", "ar-mda-kb2-luc-antagonist-p1_via")
plot_list <- list()

for(i in 1:3){
  ds_curve_i <-  ds_curves %>% 
    filter(sample_id==chems[i] & readout==outs[i]) %>%
    mutate(batch = as.factor(batch))
  plot_i <- ds_curve_i %>% 
    ggplot(aes(x=concs, y=resps, col = batch, group = batch))+
    geom_point()+
    geom_smooth(formula = y~x, method = "loess")+
    labs(x = unique(ds_curve_i$sample_name), y = outs[i])
  plot_list[[i]] <- plot_i
}

# plot_list[[1]]
ggpubr::ggarrange(plotlist = plot_list, nrow = 1, common.legend = T)
```



```{r, eval = FALSE}
# load data and get final test results
# if saved as multiple R data
output_names <- list.files(here(""))
shapes <- c("increase", "decrease", "convex", "concave")
## containers
final_pvals <- list()
final_shape <- list()
for(k in seq_along(output_names)){
  load(here(paste0("Output/", output_names[k])))
  # median p-values
  df_medp <- output_chem_k %>% group_by(Readout) %>% summarise_at(shapes, median)
  df_medp <- df_medp %>% mutate(chemical = substr(output_names[k], 1, 15))
  final_pvals[[k]] <- df_medp
  
  # shape
  df_shape <- df_medp %>%
    mutate_at(shapes, function(x){x<0.05})
  df_shape$shape <- apply(df_shape[, shapes], 1, 
                          function(x){paste(shapes[x], collapse = "-")})
  final_shape[[k]] <- df_shape
  
}

# table of p values
df_pvals <- bind_rows(final_pvals) %>%
  left_join(sample_info, by = "chemical") %>%
  mutate(Readout=gsub("mixture-","", Readout))
```


```{r}
# if all median p value already saved as csv
df_pvals <- read.csv("../Data/allpvals.csv")

# shape label
shapes <- c("increase", "decrease", "convex", "concave")
df_shape <- df_pvals %>% mutate_at(shapes, function(x){x<0.05})
df_shape$shape <- apply(df_shape[, shapes], 1, 
                          function(x){paste(shapes[x], collapse = "-")})
# category of shapes
df_shape <- df_shape %>% mutate(
  shape2 = fct_collapse(shape,
                        "Hormetic" = c("decrease-concave", "increase-convex"), 
                        "Monotone" = c("decrease", "increase"), 
                        "U shape" = c("convex", "concave"),
                        "No shape detected" = c(""),
                        other_level = "Uncertain")) %>% 
  mutate(shape2 = fct_relevel(shape2, "Hormetic", after = 0)) %>% 
  mutate(shape2 = fct_relevel(shape2, "No shape detected", after = Inf))

# remove channel 1 and 2
df_shape <- df_shape %>%
  filter(!str_ends(Readout, "ch1") & !str_ends(Readout, "ch2")) %>% 
  arrange(chemical_name) %>% 
  mutate(chemical = factor(chemical, levels = unique(chemical)))
# length(unique(df_shape$chemical)) # 90 chemicals
# length(unique(df_shape$Readout)) # 54 readouts

# add sample info
df_shape <- df_shape %>% 
  rename(sample_id = chemical) %>%
  left_join(sample_info)
```


# Overall summarization

```{r overall_sum, fig.height=20, fig.width=15}
# counts
table(df_shape$shape2)

# format so it looks nice
# plot by chemical names and 
df_shape %>%
  # mutate(shape2 = f)
  ggplot(aes(x=Readout, y = sample_id, fill=shape2))+
  geom_tile()+
  # theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 20),
  #       axis.text.y = element_text(size = 20),
  #       legend.text = element_text(size = 20),
  #       legend.title = element_blank(),
        # legend.position = "top")+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title = element_text(size=20),
        legend.title = element_blank(),
        legend.text = element_text(size = 20),
        legend.position = "top")+
  scale_x_discrete(breaks = df_shape$chemical, labels = df_shape$sample_name)+
  labs(x="Readout", y= "Chemical", fill = "Shape")
```



```{r overall_sum_label, fig.height=20, fig.width=15}
# plot by botanical group
df_shape %>%
  # group assays by channel 
  mutate(out_ch = sub(".*_", "", Readout), .after = "Readout") %>%
  mutate(out_ch = ifelse(str_ends(out_ch, "h"), "time", out_ch)) %>%
  ggplot(aes(x=Readout, y = sample_id, fill=shape2))+
  geom_tile()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 10),
        axis.text = element_text(size = 10),
        # strip.text.y = element_text(angle = 0, size = 20), 
        strip.text.y = element_text(angle = 0, size = 10),
        strip.text.x = element_text(size = 10),
        plot.margin = margin(10, 20, 10, 20),
        legend.title = element_blank(),
        legend.text = element_text(size = 10),
        legend.position = "top")+
  scale_y_discrete(breaks = df_shape$sample_id, labels = df_shape$sample_name)+
  scale_fill_manual(values = my_cols)+
  facet_grid(rows = vars(str_wrap(botanical_group)), cols = vars(out_ch), 
             scales = "free", space = "free")+
  labs(x=" ", y= " ")
```


## Scatterplot example of one assay

```{r scatter_exp, fig.height=8, fig.width=10}
df_plot <- df_pvals %>% filter(Readout == "are-bla-p1_ratio") %>%
  rename(sample_id = chemical) %>%
  left_join(sample_info)

SharpScatter(df_plot$increase, df_plot$decrease, df_plot$concave, df_plot$convex, 
             col = df_plot$botanical_group)
```

# Hormesis curves

```{r hormesis_sum,  fig.height=20, fig.width=19}
# sum(df_shape$shape2=="Hormetic")
ds_horm <- df_shape %>% 
  filter(shape2 == "Hormetic") %>% 
  select(sample_id, sample_name, Readout)

# distinguish two goldenseals
df_shape %>% 
  filter(sample_id %in% ds_horm$sample_id & Readout %in% ds_horm$Readout) %>%
  ggplot(aes(y=sample_name, x=Readout, fill=shape2))+
  geom_tile()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
        strip.text.y = element_text(angle=0, size = 15))+
  labs(y="Chemical", x= "Readout", fill = "Shape")+
  facet_grid(rows = vars(botanical_group), scales="free_y", space = "free_y")
```

## Table summary by botanical group (chemical) 

```{r}
# hormesis summary table by chemical groups
# table(df_shape$shape2)
df_shape %>% select(botanical_group, sample_name, Readout, shape2) %>% 
  group_by(botanical_group) %>% 
  summarise(nhorm = sum(shape2=="Hormetic"),
            ncurve = n(),
            pct = round(100*(nhorm/ncurve), 2)) %>%
  mutate('N (pct)' = paste0(nhorm, " (", pct, "%)")) %>%
  arrange(desc(nhorm)) %>%
  select(botanical_group, 'N (pct)') %>%
  rename("Botanical Group"=botanical_group) %>%
  kableExtra::kable(align = "lc",
                    caption = "Number of dose-response curves identified as hormesis, and percentage of hormesis curves out of all dose-reponse curves in each botanical group.") %>% 
  kableExtra::kable_styling(full_width = F)
  
```

## Table summary by assays

```{r}
# ds_horm %>% distinct(.)
# correspond each assay with their readouts
channel_assay_list <- ds_horm %>% 
  select(Readout) %>%
  distinct(.) %>%
  separate(Readout, into = c("Assay", "Channel"), sep = "_", remove = F) %>%
  group_by(Channel)
channel_assay_list <- split(channel_assay_list$Assay, f = channel_assay_list$Channel)

df_shape <- df_shape %>% 
  separate(Readout, into = c("Assay", "Channel"), sep = "_", remove = F)
```


```{r, message=FALSE}
# luc and viability 
df_shape %>% group_by(Assay, Channel) %>% 
  filter(Assay %in% channel_assay_list$luc & Assay %in% channel_assay_list$via) %>%
  group_by(Assay, Channel) %>%
  summarise(nhorm = sum(shape2=="Hormetic"),
            ncurve = n(),
            pct = round(100*(nhorm/ncurve), 2)) %>%
  mutate('N (pct)' = paste0(nhorm, " (", pct, "%)")) %>%
  select(Assay, Channel, 'N (pct)') %>%
  pivot_wider(id_cols = "Assay", values_from ='N (pct)', names_from = "Channel") %>% 
  knitr::kable(caption = "Number of dose-response curves identified as hormesis, and percentage of hormesis curves out of all dose-reponse curves of each assay in each channel.") %>%
  kableExtra::kable_styling(full_width = F)
  
```
```{r, message=FALSE}
# ratio and viability
df_shape %>% 
  filter(Assay %in% channel_assay_list$ratio & Assay %in% channel_assay_list$via) %>%
  group_by(Assay, Channel) %>%
  summarise(nhorm = sum(shape2=="Hormetic"),
            ncurve = n(),
            pct = round(100*(nhorm/ncurve), 2)) %>%
  mutate('N (pct)' = paste0(nhorm, " (", pct, "%)")) %>%
  select(Assay, Channel, 'N (pct)') %>%
  pivot_wider(id_cols = "Assay", values_from ='N (pct)', names_from = "Channel") %>% 
  knitr::kable(caption = "Number of dose-response curves identified as hormesis, and percentage of hormesis curves out of all dose-reponse curves of each assay in each channel.") %>%
  kableExtra::kable_styling(full_width = F)
```

```{r, message=FALSE}
# time
df_shape %>% group_by(Assay, Channel) %>% 
  filter(Assay %in% channel_assay_list$`0h`) %>%
  group_by(Assay, Channel) %>%
  summarise(nhorm = sum(shape2=="Hormetic"),
            ncurve = n(),
            pct = round(100*(nhorm/ncurve), 2)) %>%
  mutate('N (pct)' = paste0(nhorm, " (", pct, "%)")) %>%
  select(Assay, Channel, 'N (pct)') %>%
  pivot_wider(id_cols = "Assay", values_from ='N (pct)', names_from = "Channel") %>% 
  knitr::kable(caption = "Number of dose-response curves identified as hormesis, and percentage of hormesis curves out of all dose-reponse curves of each assay at each time.") %>%
  kableExtra::kable_styling(full_width = F)
```


## Individual dose-response curves examples

```{r ds_hormesis_exp, fig.height=10, fig.width=15}
# randomly select 6 pairs
rid <- sample(1:nrow(ds_horm), size = 15)

curve_list <- list()

for(i in 1:15){
  r <- rid[i]
  df_r <- ds_curves[ds_curves$sample_id == ds_horm$sample_id[r] &
                      ds_curves$readout == ds_horm$Readout[r], ]
  plot_r <- df_r %>% ggplot(aes(x=concs, y=resps))+
    geom_point(aes(col=as.factor(batch)), size = 2)+
    geom_smooth(method = "loess", formula = 'y ~ x', span = 0.5)+
    labs(x=ds_horm$sample_name[r], y=ds_horm$Readout[r])+
    theme(legend.position = "none")
  curve_list[[i]] <- plot_r
  
}

hcurves <- do.call(gridExtra::grid.arrange, c(curve_list, ncol = 5, nrow = 3))
# ggsave(filename = "Images/HormesisCurves.jpeg", hcurves, height = 10, width = 15)
```



# Specific assay and chemicals as examples

```{r chem_exp}
exp_chem <- sample_info$sample_id[sample_info$sample_name=="Turmeric (lot 90H73)"]
exp_assay <- c("ar-mda-kb2-luc-antagonist-p1_luc", "are-bla-p1_ratio", "rt-viability-hek293-flor-p1_40h", 
               "rt-viability-hek293-flor-p1_8h")

ds_curves %>% 
  filter(sample_id==exp_chem & readout %in% exp_assay) %>%
  ggplot(aes(x=concs, y=resps))+
    geom_point(aes(col=as.factor(batch)), size = 2)+
    geom_smooth(method = "loess", formula = 'y ~ x', span = 0.5)+
    labs(x="Turmeric (lot 90H73)", y="")+
    theme(legend.position = "none")+
  facet_wrap(~readout, scales="free_y")
  

```


```{r assay_exp}
exp_chem <- c("Ginkgo Biloba Extract (lot 020703)", "Ginkgo Biloba Extract (lot 001003)",
              "Quercetin (lot 969-0483-18BL)", "Quercetin (lot 049K1532)")
exp_assay <- "are-bla-p1_ratio"

ds_curves %>% 
  filter(readout==exp_assay) %>%
  filter(sample_id %in% unique(sample_info$sample_id[sample_info$sample_name %in% exp_chem])) %>%
  left_join(sample_info[ ,c("sample_id", "sample_name")]) %>%
  ggplot(aes(x=concs, y=resps))+
    geom_point(aes(col=as.factor(batch)), size = 2)+
    geom_smooth(method = "loess", formula = 'y ~ x', span = 0.5)+
    labs(x=" ", y="are-bla-p1_ratio")+
    theme(legend.position = "none")+
  facet_wrap(~sample_name, scales="free")
  

```

# Output from voting mechanism

## Scatter plots for p values
```{r, fig.height=6, fig.width=10}
output_names <- list.files("../Data/VoteOutput")

# random select one chemical
chem <- sample(output_names, 1)
load(paste0("../Data/VoteOutput/", chem))

df_plot <- output_chem_k %>%
  filter(!str_ends(Readout, "ch1") & !str_ends(Readout, "ch2"))
# length(unique(df_plot$Readout))
# table(df_plot$Readout)

# select a subset of readouts
select_out <- sample(unique(df_plot$Readout), 10)
df_plot <- df_plot %>% filter(Readout %in% select_out)

SharpScatter(df_plot$increase, df_plot$decrease,
             df_plot$concave,df_plot$convex, 
             col = df_plot$Readout, size_point = 0.5)
# ggsave("../Manuscripts/ReportFigs/rep_pvals_scatter.jpeg", height = 6, width = 10)
```

```{r compare}
pvals_out <- df_shape %>% select(sample_id, Readout, shape2) %>% 
  rename(shape_label = shape2, readout=Readout) %>% 
  mutate(shape_label = as.character(shape_label))
# class(pvals_out$shape_label)
# table(pvals_out$shape_label)
# table(vote_out$shape_label)

pvals_out$shape_label[pvals_out$shape_label=="Hormetic"] <- "Hormesis"

df_shape_vote <- read.csv("../Data/AllShapeVote.csv")
vote_out <- df_shape_vote %>% select(sample_id, readout, shape_label)
vote_out$readout <- gsub("mixture-","", vote_out$readout)

anti_join(pvals_out, vote_out) %>% 
  left_join(vote_out, by = c("sample_id", "readout")) %>%
  rename(median = shape_label.x,
         vote = shape_label.y) %>%
  select(median, vote) %>% table()
```
# Demonstration of how to use the scatterplot

I will directly use the 21 repitition p values to make this plot, as it essentially has no different from the 20 repitition one. Below are what I intend to plot

## Chemical: Annatto Oil Soluble (NCGC00372615-01)

- Ambiguous: mixture-rt-viability-hepg2-glo-p1_32h
- Hormesis, increase & convex: mixture-ar-mda-kb2-luc-agonist-p1_luc
- Hormesis, decrease & concave: mixture-aromatase-er-antagonist-p1_luc
- Ushape, convex: ahr-p1_luc
- Ushape, concave: mmp-p1_ratio
- Monotone, increasing: Annatto Oil Soluble (NCGC00372615-01) vs pparg-bla-agonist-p1_ratio
- Monotone, decreasing: mixture-hse-bla-p1_ratio
- Uncertain:  mixture-rt-viability-hepg2-glo-p1_8h


```{r scatter_demo, fig.height=6, fig.width=10}
load(here("Data/VoteOutput/NCGC00372615-01.RData"))
plot_outs <- c("rt-viability-hepg2-glo-p1_32h", "rt-viability-hepg2-glo-p1_8h", 
               "ar-mda-kb2-luc-agonist-p1_luc", "aromatase-er-antagonist-p1_luc", 
               "ahr-p1_luc" ,"mmp-p1_ratio", 
               "pparg-bla-agonist-p1_ratio", "hse-bla-p1_ratio")

# plot_outs <- df_shape %>% filter(sample_id == "NCGC00372615-01" & shape2 == "U shape")
# plot_outs <- plot_outs$Readout

color_label <- c("Ambiguous", "No shape", 
                 "Hormesis (convex-increase) ", "Hormesis (concave-decrease)",
                 "U shape (convex)", "U shape (concave)", 
                 "Monotone (increase)", "Monotone (decrease)")
names(color_label) <- plot_outs

df_plot <- output_chem_k %>%
  mutate(Readout = gsub("mixture-", "", Readout)) %>%
  filter(Readout %in% plot_outs)
SharpScatter(df_plot$increase, df_plot$decrease,
             df_plot$concave,df_plot$convex, 
             col = df_plot$Readout, size_point = 2)+
  scale_color_discrete("", labels = color_label)
```


```{r demo_curves, fig.height=6, fig.width=10}
ds_curves %>%
  filter(sample_id=="NCGC00372615-01" & readout %in% plot_outs) %>%
    ggplot(aes(x=concs, y=resps))+
    geom_point(aes(col = readout))+
    geom_smooth(aes(col = readout), span = 0.7)+
    theme(legend.position = "none")+
    facet_wrap(~readout, scales="free", nrow = 2)

```

## Chemical: Turmeric (lot 90H73) (NCGC00372872-01)

- ambigous: aromatase-er-aromatase-p1_luc
- no shape: elg1-luc-agonist-p1_luc
- hormesis, convex-increase: rt-viability-hek293-flor-p1_0h
- hormesis, concave-decrease: rt-viability-hepg2-glo-p1_0h
- U shape, concave: rt-viability-hek293-glo-p1_0h
- U shape, convex: 
- Monotone, increase: 
- Monoton decrease: mmp-p1_ratio

```{r scatter_demo2, fig.height=6, fig.width=10}
load(here("Data/VoteOutput/NCGC00372872-01.RData"))
# examination
# plot_outs <- df_shape %>% filter(sample_id == "NCGC00372872-01" & shape2 == "U shape")
# plot_outs <- plot_outs$Readout
plot_outs <- c("aromatase-er-aromatase-p1_luc", "elg1-luc-agonist-p1_luc",
               "rt-viability-hek293-flor-p1_0h" ,"rt-viability-hepg2-glo-p1_0h",
               "rt-viability-hek293-glo-p1_0h", "mmp-p1_ratio")
color_label <- c("Ambiguous", "No shape", 
                 "Hormesis (convex-increase) ", "Hormesis (concave-decrease)",
                 "U shape (concave) & ambiguous", 
                 "Monotone (decrease)")
names(color_label) <- plot_outs

df_plot <- output_chem_k %>%
  mutate(Readout = gsub("mixture-", "", Readout)) %>%
  filter(Readout %in% plot_outs) 

SharpScatter(df_plot$increase, df_plot$decrease,
             df_plot$concave,df_plot$convex, 
             col = df_plot$Readout, size_point = 2)+
  scale_color_discrete("", labels = color_label)
```


## Chemical: Kava kava extract (lot 562) (NCGC00372872-01)

- ambigous: rt-viability-hepg2-flor-p1_40h
- no shape: rt-viability-hepg2-flor-p1_8h
- hormesis, convex-increase: elg1-luc-agonist-p1_luc
- hormesis, concave-decrease: mmp-p1_ratio
- U shape, concave: rt-viability-hek293-glo-p1_16h
- U shape, convex: p53-bla-p1_ratio
- Monotone, increase: er-luc-bg1-4e2-agonist-p1_luc
- Monoton decrease: 

```{r scatter_demo3, fig.height=6, fig.width=10}
load(here("Data/VoteOutput/NCGC00372762-01.RData"))
# examination
# plot_outs <- df_shape %>% filter(sample_id == "NCGC00372762-01") %>% 
#   # select(shape) %>% table()
#   filter(shape == "decrease")
# plot_outs <- plot_outs$Readout

plot_outs <- c("rt-viability-hepg2-flor-p1_40h", "rt-viability-hepg2-flor-p1_8h",
               "elg1-luc-agonist-p1_luc", "mmp-p1_ratio",
               "rt-viability-hek293-glo-p1_16h", "p53-bla-p1_ratio",
               "er-luc-bg1-4e2-agonist-p1_luc")
color_label <- c("Ambiguous", "No shape", 
                 "Hormesis (convex-increase) ", "Hormesis (concave-decrease)",
                 "U shape (concave)", "U shape (convex)", 
                 "Monotone (increase)")
names(color_label) <- plot_outs

df_plot <- output_chem_k %>%
  mutate(Readout = gsub("mixture-", "", Readout)) %>%
  filter(Readout %in% plot_outs) 

SharpScatter(df_plot$increase, df_plot$decrease,
             df_plot$concave,df_plot$convex, 
             col = df_plot$Readout, size_point = 2)+
  scale_color_discrete("", labels = color_label)
```



```{r demo_curves3, fig.height=6, fig.width=10}
ds_curves %>%
  filter(sample_id=="NCGC00372762-01" & readout %in% plot_outs) %>%
    ggplot(aes(x=concs, y=resps))+
    geom_point(aes(col = readout))+
    geom_smooth(aes(col = readout), span = 0.7)+
    theme(legend.position = "none")+
    facet_wrap(~readout, scales="free", nrow = 2)

```
